Studying the Performance of the Jellyfish Search Optimiser for the Application of Projection Pursuit

### 2024-06-30 optim.qmd
update illustration, small edits

---
title: Studying the Performance of the Jellyfish Search Optimiser for the Application of Projection Pursuit
header-includes:
   - \usepackage{algorithm}
author:
  - name: H. Sherry Zhang
    email: huize.zhang@austin.utexas.edu
    affiliations: 
        - id: 1
          name: University of Texas at Austin
          department: Department of Statistics and Data Sciences
          city: Austin
          country: United States
          postal-code: 78751
    attributes:
        corresponding: true
  - name: Dianne Cook
    email: dicook@monash.edu
    affiliations:
        - id: 2
          name: Monash University
          department: Department of Econometrics and Business Statistics
          city: Melbourne
          country: Australia
          postal-code: 3800
  - name: Nicolas  Langrené
    email: nicolaslangrene@uic.edu.cn
    affiliations:
        - id: 3
          name: Guangdong Provincial Key Laboratory of Interdisciplinary Research and Application for Data Science, BNU-HKBU United International College
          department: Department of Mathematical Sciences
          city: Zhuhai
          country: China
          postal-code: 519087
  - name: Jessica Wai Yin Leung
    email: Jessica.Leung@monash.edu
    affiliations:
        - id: 2
          name: Monash University
          department: Department of Econometrics and Business Statistics
          city: Melbourne
          country: Australia
          postal-code: 3800
abstract: |
  The projection pursuit (PP) guided tour interactively optimises a criteria function known as the PP index, to explore high-dimensional data by revealing interesting projections. The optimisation in PP can be non-trivial, involving non-smooth functions and optima with small a "squint angle", detectable only from close proximity. To address these challenges, this study investigates the performance of a recently introduced swarm-based algorithm, Jellyfish Search Optimiser (JSO), for optimising PP indexes. The performance of JSO for visualising data is evaluated across various hyper-parameter settings and compared with existing optimisers. Additionally, this work proposes novel methods to quantify two properties of the PP index -- smoothness and squintability -- that capture the complexities inherent in PP optimisation problems. These two metrics are evaluated with JSO hyper-parameters to determine their effect on JSO success rate. The JSO algorithm has been implemented in the `tourr` package, while functions to calculate  smoothness and squintability of indexes are available in the `ferrn` package. 
keywords: 
  - projection pursuit
  - jellyfish search optimiser (JSO)
  - optimisation
  - grand tour
  - high-dimensional data
  - exploratory data analysis
date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      name: Journal of Multivariate Analysis
      formatting: preprint
      model: 3p
      cite-style: numbername
editor: 
  markdown: 
    wrap: 72
crossref: 
  eq-prefix: ""
---

@@

# Introduction \[Nicolas and Jessica\]

@@

The paper is structured as follows. @sec-background introduces the background of PP guided tour and reviews existing optimisers and index functions in the literature. @sec-theory details JSO and introduces metrics that measure different properties of PP indexes. @sec-sim-deets outlines the data structures, numerical summaries, and visualisations used in the simulation studies conducted in @sec-app to assess JSO's performance. @sec-sim-res presents two sets of simulation experiments, comparing JSO permformance improvements relative to an existing optimiser, and studies the impact of different PP index properties on the optimisation performance. @sec-conclusion summarises the work and provides suggestions for future directions.

# Projection pursuit, tours, index functions and optimisation \[Di and Sherry\] {#sec-background}

@@

# The jellyfish optimiser and properties of PP indexes \[Nicolas and Jessica\] {#sec-theory}

@@

# Results \[Di and Sherry\] {#sec-sim-res}

@@

# Conclusion \[Di and Sherry\] {#sec-conclusion}

@@

>>>>

---
title: Studying the Performance of the Jellyfish Search Optimiser for the Application of Projection Pursuit
header-includes:
   - \usepackage{algorithm}
author:
  - name: H. Sherry Zhang
    email: huize.zhang@austin.utexas.edu
    affiliations: 
        - id: 1
          name: University of Texas at Austin
          department: Department of Statistics and Data Sciences
          city: Austin
          country: United States
          postal-code: 78751
    attributes:
        corresponding: true
  - name: Dianne Cook
    email: dicook@monash.edu
    affiliations:
        - id: 2
          name: Monash University
          department: Department of Econometrics and Business Statistics
          city: Melbourne
          country: Australia
          postal-code: 3800
  - name: Nicolas  Langrené
    email: nicolaslangrene@uic.edu.cn
    affiliations:
        - id: 3
          name: Guangdong Provincial Key Laboratory of Interdisciplinary Research and Application for Data Science, BNU-HKBU United International College
          department: Department of Mathematical Sciences
          city: Zhuhai
          country: China
          postal-code: 519087
  - name: Jessica Wai Yin Leung
    email: Jessica.Leung@monash.edu
    affiliations:
        - id: 2
          name: Monash University
          department: Department of Econometrics and Business Statistics
          city: Melbourne
          country: Australia
          postal-code: 3800
abstract: |
  The projection pursuit (PP) guided tour interactively optimises a criteria function known as the PP index, to explore high-dimensional data by revealing interesting projections. The optimisation in PP can be non-trivial, involving non-smooth functions and optima with small a "squint angle", detectable only from close proximity. To address these challenges, this study investigates the performance of a recently introduced swarm-based algorithm, Jellyfish Search Optimiser (JSO), for optimising PP indexes. The performance of JSO for visualising data is evaluated across various hyper-parameter settings and compared with existing optimisers. Additionally, this work proposes novel methods to quantify two properties of the PP index -- smoothness and squintability -- that capture the complexities inherent in PP optimisation problems. These two metrics are evaluated with JSO hyper-parameters to determine their effects on JSO success rate. The JSO algorithm has been implemented in the `tourr` package and functions to calculate  smoothness and squintability of indexes are available in the `ferrn` package. 
keywords: 
  - projection pursuit
  - jellyfish search optimiser (JSO)
  - optimisation
  - grand tour
  - high-dimensional data
  - exploratory data analysis
date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      name: Journal of Multivariate Analysis
      formatting: preprint
      model: 3p
      cite-style: numbername
editor: 
  markdown: 
    wrap: 72
crossref: 
  eq-prefix: ""
---

@@

# Introduction

@@

The paper is structured as follows. @sec-background introduces the background of PP guided tour and reviews existing optimisers and index functions in the literature. @sec-theory details JSO and introduces metrics that measure different properties of PP indexes. @sec-sim-deets details two simulation experiments to assess JSO's performance: one comparing JSO's permformance improvements relative to an existing optimiser, Creeping Random Search (CRS), and the other studying the impact of different PP index properties on optimisation performance. @sec-sim-res presents the results. @sec-conclusion summarises the work and provides suggestions for future directions.

# Projection pursuit, tours, index functions and optimisation {#sec-background}

@@

# The jellyfish optimiser and properties of PP indexes {#sec-theory}

@@

# Results {#sec-sim-res}

@@

# Conclusion {#sec-conclusion}

@@

# Acknowledgement

The article is created using Quarto [@Allaire_Quarto_2022] in R [@R].
The source code for reproducing the work reported in this paper can be found at: <https://github.com/huizezhang-sherry/paper-optim>. Nicolas Langrené acknowledges the partial support of the Guangdong Provincial Key Laboratory IRADS (2022B1212010006, R0400001-22) and the UIC Start-up Research Fund UICR0700041-22.

### 2024-06-30 optim.tex
update illustration, small edits

\begin{abstract}
The projection pursuit (PP) guided tour interactively optimises a
criteria function known as the PP index, to explore high-dimensional
data by revealing interesting projections. The optimisation in PP can be
non-trivial, involving non-smooth functions and optima with small a
``squint angle'', detectable only from close proximity. To address these
challenges, this study investigates the performance of a recently
introduced swarm-based algorithm, Jellyfish Search Optimiser (JSO), for
optimising PP indexes. The performance of JSO for visualising data is
evaluated across various hyper-parameter settings and compared with
existing optimisers. Additionally, this work proposes novel methods to
quantify two properties of the PP index -- smoothness and squintability
-- that capture the complexities inherent in PP optimisation problems.
These two metrics are evaluated with JSO hyper-parameters to determine
their effect on JSO success rate. The JSO algorithm has been implemented
in the \texttt{tourr} package, while functions to calculate smoothness
and squintability of indexes are available in the \texttt{ferrn}
package.
\end{abstract}

@@

\section{Introduction {[}Nicolas and
Jessica{]}}\label{introduction-nicolas-and-jessica}

@@

The paper is structured as follows. Section~\ref{sec-background}
introduces the background of PP guided tour and reviews existing
optimisers and index functions in the literature.
Section~\ref{sec-theory} details JSO and introduces metrics that measure
different properties of PP indexes. Section~\ref{sec-sim-deets} outlines
the data structures, numerical summaries, and visualisations used in the
simulation studies conducted in \textbf{?@sec-app} to assess JSO's
performance. Section~\ref{sec-sim-res} presents two sets of simulation
experiments, comparing JSO permformance improvements relative to an
existing optimiser, and studies the impact of different PP index
properties on the optimisation performance. Section~\ref{sec-conclusion}
summarises the work and provides suggestions for future directions.

\section{Projection pursuit, tours, index functions and optimisation
{[}Di and Sherry{]}}\label{sec-background}

@@

\section{The jellyfish optimiser and properties of PP indexes {[}Nicolas
and Jessica{]}}\label{sec-theory}

@@

\begin{tcolorbox}[enhanced jigsaw, toptitle=1mm, bottomtitle=1mm, bottomrule=.15mm, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white, titlerule=0mm, leftrule=.75mm, coltitle=black, opacitybacktitle=0.6, title={Algorithm: Jellyfish Optimizer Pseudo Code}, breakable, opacityback=0, colback=white, left=2mm, toprule=.15mm, arc=.35mm, colframe=quarto-callout-note-color-frame]

@@

\includegraphics[width=4.49in,height=\textheight]{figures/success-rate.png}

@@

\section{Results {[}Di and Sherry{]}}\label{sec-sim-res}

@@

\section{Conclusion {[}Di and Sherry{]}}\label{sec-conclusion}

@@

>>>>

\begin{abstract}
The projection pursuit (PP) guided tour interactively optimises a
criteria function known as the PP index, to explore high-dimensional
data by revealing interesting projections. The optimisation in PP can be
non-trivial, involving non-smooth functions and optima with small a
``squint angle'', detectable only from close proximity. To address these
challenges, this study investigates the performance of a recently
introduced swarm-based algorithm, Jellyfish Search Optimiser (JSO), for
optimising PP indexes. The performance of JSO for visualising data is
evaluated across various hyper-parameter settings and compared with
existing optimisers. Additionally, this work proposes novel methods to
quantify two properties of the PP index -- smoothness and squintability
-- that capture the complexities inherent in PP optimisation problems.
These two metrics are evaluated with JSO hyper-parameters to determine
their effects on JSO success rate. The JSO algorithm has been
implemented in the \texttt{tourr} package and functions to calculate
smoothness and squintability of indexes are available in the
\texttt{ferrn} package.
\end{abstract}

@@

\section{Introduction}\label{introduction}

@@

The paper is structured as follows. Section~\ref{sec-background}
introduces the background of PP guided tour and reviews existing
optimisers and index functions in the literature.
Section~\ref{sec-theory} details JSO and introduces metrics that measure
different properties of PP indexes. Section~\ref{sec-sim-deets} details
two simulation experiments to assess JSO's performance: one comparing
JSO's permformance improvements relative to an existing optimiser,
Creeping Random Search (CRS), and the other studying the impact of
different PP index properties on optimisation performance.
Section~\ref{sec-sim-res} presents the results.
Section~\ref{sec-conclusion} summarises the work and provides
suggestions for future directions.

\section{Projection pursuit, tours, index functions and
optimisation}\label{sec-background}

@@

\section{The jellyfish optimiser and properties of PP
indexes}\label{sec-theory}

@@

\begin{tcolorbox}[enhanced jigsaw, titlerule=0mm, opacitybacktitle=0.6, colframe=quarto-callout-note-color-frame, colbacktitle=quarto-callout-note-color!10!white, toptitle=1mm, arc=.35mm, colback=white, title={Algorithm: Jellyfish Optimizer Pseudo Code}, bottomtitle=1mm, leftrule=.75mm, coltitle=black, bottomrule=.15mm, toprule=.15mm, opacityback=0, left=2mm, rightrule=.15mm, breakable]

@@

\includegraphics[width=4.52in,height=\textheight]{figures/success-rate.png}

@@

\section{Results}\label{sec-sim-res}

@@

\section{Conclusion}\label{sec-conclusion}

@@

\section{Acknowledgement}\label{acknowledgement}

The article is created using Quarto \citep{Allaire_Quarto_2022} in R
\citep{R}. The source code for reproducing the work reported in this
paper can be found at:
\url{https://github.com/huizezhang-sherry/paper-optim}. Nicolas Langrené
acknowledges the partial support of the Guangdong Provincial Key
Laboratory IRADS (2022B1212010006, R0400001-22) and the UIC Start-up
Research Fund UICR0700041-22.
